A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-02-09 paper4.Rmd
fix details and short tables.

The impact of ambient temperature variations on infant mortality is studied for 
a northern Sweden coastal area, the Umeå and Skellefteå regions, during the 
first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
neonatal mortality and temperature variations in a larger geographical area containing 
the present one during the 
years 1880--1950. Climate and mortality in general is a research area that has 
generated great interest over the last years, see @tbgb10.

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with sex, social class, and legitimacy. Studies are performed 
separately for neonatal  and postneonatal mortality, and for winter and summer seasons, 
and the classification into endogenous and exogenous factors will be discussed.

@@

**socStatus** Social status of family, based on HISCLASS.

@@

Figure \@ref(fig:mortstat) shows the average monthly crude infant mortality, and a clear seasonal pattern is visible.

@@

The average monthly neonatal mortality is shown in Figure \@ref(fig:neodata).

@@

The average monthly postneonatal mortality is shown in Figure \@ref(fig:pneodata).

@@

@@

There are three measurements per day, or 21 per week. In the forthcoming analyses,
the weekly data are summarized in a few measurements, see Table \@ref(tab:cond).

@@

```{r shownonprop, cache = FALSE}
tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
```

From Table \@ref(tab:shownonprop) we can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
*infant mortality*, we can say that these caovariates have non-proportional effects. As an example, see Figure

@@

```{r neoanalw, results = "asis", cache = FALSE}
ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanalw", keep = NULL)
```

@@

```{r neoanals, results = 'asis', cache = FALSE}
ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", 
     label = "tab:neoanals", keep = NULL)
```

@@

```{r postanalw, results='asis', cache = FALSE}
tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal mortality.", rownames = TRUE)
##ltx2(fit.tpch, dr = dr, caption = "Postneonatal mortality, October to March. Adjusted for social branch, sex, ##illegitimacy, parity, time period, and subregion.", label = "tab:postanalw", keep = NULL)
```

The result in Table \@ref(tab:postanalw) shows that the temperature effects vary with time period,
so we continue with seaparate analyses for the time periods `r levels(tabpost$period)`.

```{r setformw}
form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socStatus + subreg + sex + parity")
```

@@

```{r postw1, results = "asis", cache = FALSE}
ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw1", keep = NULL)
```

@@

```{r postw2, results = "asis", cache = FALSE}
ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw2", keep = NULL)
```

@@

```{r postw3, results = "asis", cache = FALSE}
ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for social branch, sex, illegitimacy, parity, and subregion.", label = "tab:postw3", keep = NULL)
```

@@

```{r postanals, results='asis', cache = FALSE}
ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for social branch, sex, illegitimacy, parity, time period, and subregion.", label = "tab:postanals", keep = NULL)
```

The result in Table \@ref(tab:postanals) shows that *climate* (`emintemp`) is more important than 
*weather* (`excessTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
covariates (not shown).

>>>>

The impact of ambient temperature variations on infant mortality is studied for 
two northern Sweden areas, the Umeå coastal region and Skellefteå coastal and inland regions, during the 
first half of the twentieth century. Two recent papers [@junkka; @karlsson] studied 
neonatal mortality and temperature variations in this geographical area during the 
years 1880--1950. Climate and mortality in general is a research area that has 
generated great interest over the last years, see @tbgb10 for an example of our own efforts.

The effect of seasonal variation and the occurrence of extreme monthly temperatures
is studied and interacted with social class and time period. Studies are performed 
separately for neonatal  and postneonatal mortality, and for winter and summer seasons.

@@

@@

Figure \@ref(fig:mortstat) shows the average weekly crude infant mortality, and a clear seasonal pattern is visible.

@@

The average weekly neonatal mortality is shown in Figure \@ref(fig:neodata).

@@

The average weekly postneonatal mortality is shown in Figure \@ref(fig:pneodata).

@@

Note that the estimates of neonatal and postneonatal mortality are non-standard here: 
They are calculated as *occurrence-exposure rates*, that is, the number of deaths
divided by *exposure time*. They are thus crude *hazard rates*, not probabilities, which is 
more common in definitions of mortality.

@@

There are three measurements per day, or 21 per week. In the forthcoming analyses,
the weekly data are summarized in a few measurements, see Table \@ref(tab:cond).
Our rule for week numbering is that week No. 1 always start on January 1. Week No. 52
 always ends at December 31, and so will be eight days long, except for leap years, when
 it will be nine days long.

@@

Covariates may affect neonatal and postneonatal mortality differently, and one way to 
investigate that is to stratify the infant mortality data into two age intervals, one from
birth to one month of age and the other from one month to one year of age. Then the
interaction between the stratum variable and other variables of interest is investigated, 
with this result:

```{r shownonprop, cache = FALSE}
dr.all
##tbl(round(dr.all, 4), caption = "Proportionality tests, neonatal vs. postneonatal.", rownames = TRUE)
```

We can see that the variables *emeantemp*, *lowTemp*, *socStatus* and *period* all have different effects on mortality in the neonatal case compared to the postneonatal case. In terms of 
*infant mortality*, we can say that these caovariates have non-proportional effects.

@@

```{r neoanalw, results = "asis", cache = FALSE}
ltx2(fit.neow, dr = dr, caption = "Neonatal mortality, October to March.", 
     label = "tab:neoanalw", keep = NULL)
```

@@

```{r neoanals, results = 'asis', cache = FALSE}
ltx2(fit.neos, dr = dr.neos, caption = "Neonatal mortality, April to September.", 
     label = "tab:neoanals", keep = NULL)
```

@@

Interaction tests with temperature and period, postneonatal mortality:

```{r postanalw, cache = FALSE}
round(dr.postw, 4)
#tbl(round(dr.postw, 4), caption = "Interaction tests with temperature and period, postneonatal ##mortality.", rownames = TRUE)
```

This shows that the temperature effects do not vary with time period,
but we continue with separate analyses for the time periods `r levels(tabpost$period)`. Just
to make sure, will be joined when appropriate.

```{r setformw}
form <- as.formula("oe(event, exposure) ~ lowTemp + highTemp + extemp + emeantemp + socBranch + subreg + sex + parity")
```

@@

```{r postw1, results = "asis", cache = FALSE}
ltx2(fit.postw1, dr = dr.postw1, caption = "Postneonatal mortality, October to March 1895-1914. Adjusted for social branch, sex, parity, and subregion.", label = "tab:postw1", keep = 5)
```

@@

```{r postw2, results = "asis", cache = FALSE}
ltx2(fit.postw2, dr = dr.postw2, caption = "Postneonatal mortality, October to March 1914-1934. Adjusted for sex, parity, and subregion.", label = "tab:postw2", keep = 5)
```

@@

```{r postw3, results = "asis", cache = FALSE}
ltx2(fit.postw3, dr = dr.postw3, caption = "Postneonatal mortality, October to March 1935-1950. Adjusted for sex, parity, and subregion.", label = "tab:postw3", keep = 5)
```

@@

```{r postanals, results='asis', cache = FALSE}
ltx2(fit.posts, dr = dr, caption = "Postneonatal mortality, April to September. Adjusted for sex, parity, time period, and subregion.", label = "tab:postanals", keep = 5)
```

The result in Table \@ref(tab:postanals) shows that *climate* (`emeantemp`) is more important than 
*weather* (`exTemp`). Moreover, no signs of interaction between weather or climate and the rest of 
covariates (not shown).
